---
title: "Storm Data Analysis"
output: html_document
---

This is an analysis of storm and severe weather data from the NOAA database. The goal is to analyze the data and answer two questions:

1.Across the United States, which types of events are most harmful with respect to population health?

2.Across the United States, which types of events have the greatest economic consequences?

##Data Processing 
```{r}
# read in file
file <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(file, destfile = "tempdata.csv.bz2", method = "curl")
stormData <- read.csv("./tempdata.csv.bz2")

# subset relevant columns
stormData <- stormData[,c('EVTYPE','FATALITIES','INJURIES', 'PROPDMG', 'PROPDMGEXP', 'CROPDMG', 'CROPDMGEXP')]

# get property damage
stormData$PROPDMGNUM = 0
stormData[stormData$PROPDMGEXP == "H", ]$PROPDMGNUM = stormData[stormData$PROPDMGEXP == "H", ]$PROPDMG * 10^2
stormData[stormData$PROPDMGEXP == "K", ]$PROPDMGNUM = stormData[stormData$PROPDMGEXP == "K", ]$PROPDMG * 10^3
stormData[stormData$PROPDMGEXP == "M", ]$PROPDMGNUM = stormData[stormData$PROPDMGEXP == "M", ]$PROPDMG * 10^6
stormData[stormData$PROPDMGEXP == "B", ]$PROPDMGNUM = stormData[stormData$PROPDMGEXP == "B", ]$PROPDMG * 10^9

# get crop damage
stormData$CROPDMGNUM = 0
stormData[stormData$CROPDMGEXP == "H", ]$CROPDMGNUM = stormData[stormData$CROPDMGEXP == "H", ]$CROPDMG * 10^2
stormData[stormData$CROPDMGEXP == "K", ]$CROPDMGNUM = stormData[stormData$CROPDMGEXP == "K", ]$CROPDMG * 10^3
stormData[stormData$CROPDMGEXP == "M", ]$CROPDMGNUM = stormData[stormData$CROPDMGEXP == "M", ]$CROPDMG * 10^6
stormData[stormData$CROPDMGEXP == "B", ]$CROPDMGNUM = stormData[stormData$CROPDMGEXP == "B", ]$CROPDMG * 10^9
```

##Results
1.Across the United States, which types of events are most harmful with respect to population health?

We'll look at Fatalaties first:
```{r}
# get ggplot to use
library(ggplot2)

# subset and aggregate fatality data
fats <- aggregate(FATALITIES ~ EVTYPE, data=stormData, sum)
fats <- fats[order(-fats$FATALITIES), ][1:10, ]
fats$EVTYPE <- factor(fats$EVTYPE, levels = fats$EVTYPE)

# plot fatality data
ggplot(fats, aes(x = EVTYPE, y = FATALITIES)) + 
        geom_bar(stat = "identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
        xlab("Event Type") + 
        ylab("Fatalities") + 
        ggtitle("Top 10 Weather Events by Fatalaties")
```

And then injuries:
```{r}
# get ggplot to use
library(ggplot2)

# subset and aggregate injury data
inj <- aggregate(INJURIES ~ EVTYPE, data=stormData, sum)
inj <- inj[order(-inj$INJURIES), ][1:10, ]
inj$EVTYPE <- factor(inj$EVTYPE, levels = inj$EVTYPE)

# plot injuruy data
ggplot(inj, aes(x = EVTYPE, y = INJURIES)) + 
        geom_bar(stat = "identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
        xlab("Event Type") + ylab("Injuries") + 
        ggtitle("Top 10 Weather Events by Injuries")
```


Across the United States, which types of events have the greatest economic consequences?
```{r}
# get ggplot to use
library(ggplot2)

# subset and aggregate crop and property damage data
dam <- aggregate(PROPDMGNUM + CROPDMGNUM ~ EVTYPE, data=stormData, sum)
names(dam) = c("EVTYPE", "TOTALDAMAGE")
dam <- dam[order(-dam$TOTALDAMAGE), ][1:10, ]
dam$EVTYPE <- factor(dam$EVTYPE, levels = dam$EVTYPE)

# plot crop and property damage data
ggplot(dam, aes(x = EVTYPE, y = TOTALDAMAGE)) + 
      geom_bar(stat = "identity") + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
      xlab("Event Type") + ylab("Damages ($)") + 
      ggtitle("Top 10 Weather Events by Property and Crop Damage")
``` 

~fin